Under-approximation of reachability through static analysis in multivalued asynchronous networks

### 2013-03-29 parts/def-ph.tex
defs + demo

\subsection{Hypothesis} \towrite{Reformuler}

@@

Let $\PH$ be a PH model on which we want to apply the static analysis.
In this paper, we focus on a specific class of models that are equivalent to boolean networks.

@@

In \pref{def:cs} we define the notion of well-formed component.
A sort $a$ is a well-formed component iff
it is hit only by actions of priority $2$.
\begin{definition}
\label{def:component}
  A sort $a \in \PHs$ is said to be a well-formed component iff:
  \begin{itemize}
    \item $\forall h \in \PHh, \PHsort(\PHtarget(h)) = a \Rightarrow \prio(h) = 2$
  \end{itemize}
\end{definition}
We call $\components$ the set of well-formed components.

Given this first hypothesis, we define the neighbor \towrite{revoir terminologie : neighbors ? vicinity ?} sorts and actions (\pref{def:neighbors}).
\begin{definition}
\label{def:chainsactions}
  We define the set of chains of actions $\PHh(b, a)$ between two sorts $b$ and $a$ as below:
  \begin{align*}
  \PHh(b, a) = \{ (h_i)_{i \in \segm{0}{s}} \in (\PHh^{(1)})^{s+1} &\mid s \in \sN, \PHsort(\PHhitter(h_0)) = b \wedge \PHsort(\PHtarget(h_s)) = a \\
    & \wedge \forall i \in \segm{1}{s}, \PHsort(\PHtarget(h_{i-1})) = \PHsort(\PHhitter(h_{i})) \}
  \end{align*}
\end{definition}
\begin{definition}
\label{def:neighbors}
  The set of neighbour sorts $\Vs(a)$ of a given sort $a$ is the set of components that can interact with it through actions of priority $1$.
  The set of neighbour actions $\Vh(a)$ of $a$ is the set of actions of priority $1$ influencing $a$.
  $$\Vs(a) = \{ b \in \components \mid \PHh(b, a) \neq \emptyset \}$$
  $$\Vh(a) = \{ h \in \PHh^{(1)} \mid \PHsort(\PHhitter(h)) \in \Vs(a) \wedge \PHsort(\PHtarget(h)) \in \Vs(a) \cup \{ a \} \}$$
\end{definition}

@@

In \pref{def:component} we define the notion of well-formed cooperative sort.
A cooperative sort $a$ is well-formed iff
it is hit only by actions of priority $1$,
if any sub-state of its neighbourhood is represented by at least one local steady state in this sort,
and if any of its processes represents at least one sub-state of its neighbourhood.
\begin{definition}
\label{def:cs}
  A sort $a \in \PHs$ is said to be a well-formed cooperative sort iff:
  \begin{itemize}
    \item $\forall h \in \PHh, \PHsort(\PHtarget(h)) = a \Rightarrow \prio(h) = 1$
    \item $\forall s \in \underset{b \in \Vs(a)}{\times} \PHl_b, |\pfp_s(a)| = 1$
    \item $\forall a_i \in \PHl_a, \exists s \in \underset{b \in \Vs(a)}{\times} \PHl_b, a_i \in \pfp_s(a)$
  \end{itemize}
\end{definition}
We call $\cs$ the set of well-formed cooperative sorts.
%We note that: $\forall a \in \components, \Vs(a) = \emptyset \wedge \Vh(a) = \emptyset$
As any local state of neighbour sorts of a cooperative sort leads to a local steady-state, we denote $\csState(a_i)$ the set of local states represented by the process $a_i$ of the cooperative sort $a$.
\begin{definition}[$\csState$]
\label{def:csState}
  \towrite{À reformuler}
  $$\csState(a_i) = \{ s \in \underset{b \in \Vs(a)}{\times} \PHl_b \mid \pfp_{s}(a) = a_i \}$$
\end{definition}

@@

\subsection{Consequences of the restrictions}
\pref{th:hcompcomp} states that for a given state $s$ and for any action $h = \PHhit{a_i}{b_j}{b_k}$ between well-formed components, $h$ can be played after a series of hits that do not prevent it to be fired.
\pref{th:hcscomp} states the same if $a$ is a well-formed cooperative sort, under the condition that $a_i$ reflects the state of the neighborhood of $a$.
\begin{theorem}
\label{th:hcompcomp}
  $\forall s \in \PHl, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh, a,b \in \components,$\\
  $(\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j) \Rightarrow (\exists \delta \in \restriction{\Sce}{1}, \text{$h$ can be played in $s \PHplay \delta$})$
\end{theorem}
\begin{proof}
  There exists a scenario $\delta$ so that: $(s \PHplay \delta) = \update(s)$.
  As $a,b \in \components$, $a_i \in s \PHplay \delta$ and $b_j \in s \PHplay \delta$.
  Finally, by definition of $\update(s)$, no action of priority $1$ can be played in $s \PHplay \delta$, thus $h$ can be played in $s \PHplay \delta$.
\end{proof}
\begin{theorem}
\label{th:hcscomp}
  $\forall s \in \PHl, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh, a \in \cs, b \in \components$\\
  $(\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j \wedge \pfp_s(a) = a_i) \Rightarrow (\exists \delta \in \restriction{\Sce}{1}, \text{$h$ can be played in $s \PHplay \delta$})$
\end{theorem}
\begin{proof}
  There exists a scenario $\delta$ so that: $(s \PHplay \delta) = \update(s)$.
  As $a_i \in \pfp_s(a)$, $a_i \in s \PHplay \delta$.
  Furthermore, $b \in \components$, thus $b_j \in s \PHplay \delta$.
  Finally, by definition of $\update(s)$, no action of priority $1$ can be played in $s \PHplay \delta$, thus $h$ can be played in $s \PHplay \delta$.
\end{proof}

>>>>

\subsection{Hypothesis} \towrite{Reformuler}
\label{ssec:hypothesis}

@@

In the scope of this paper, we focus on a specific class of Process Hitting models.
We define here the restrictions that lead to this class of models \todo{and show that they are equivalent to discrete networks}.
%Let $\PH$ be a PH model on which we want to apply the static analysis.
%In this paper, we focus on a specific class of models that are equivalent to boolean networks.

@@

This restriction allows to models two kinds of actions: controllable actions that model the non-determinacy of biological evolutions,
and uncontrollable actions that are used to model non-biological behaviours in the model.

In \pref{def:component} we define the notion of well-formed component.
A sort $a$ is a well-formed component iff
it is hit only by actions of priority $2$.
\begin{definition}
\label{def:component}
  A sort $a \in \PHs$ is said to be a well-formed component iff:
  \begin{itemize}
    \item $\forall h \in \PHh, \PHsort(\PHtarget(h)) = a \Rightarrow \prio(h) = 2$
  \end{itemize}
\end{definition}
We call $\components$ the set of well-formed components.

\begin{definition}
\label{def:chainsactions}
  We define the set of chains of actions $\PHh(b, a)$ between two sorts $b$ and $a$ as below:
  \begin{align*}
  \PHh(b, a) = \{ (h_i)_{i \in \segm{0}{s}} \in (\PHh^{(1)})^{s+1} &\mid s \in \sN, \PHsort(\PHhitter(h_0)) = b \wedge \PHsort(\PHtarget(h_s)) = a \\
    & \wedge \forall i \in \segm{1}{s}, \PHsort(\PHtarget(h_{i-1})) = \PHsort(\PHhitter(h_{i})) \}
  \end{align*}
\end{definition}

\paragraph{Cycle-freeness between actions of priority $1$}
There is no cycle in chains of actions: $\forall a \in \PHs, \PHh(a, a) = \emptyset$.

Given this first hypothesis, we define the neighbor \towrite{revoir terminologie : neighbors ? vicinity ?} sorts and actions (\pref{def:neighbors}).
\begin{definition}
\label{def:neighbors}
  The set of neighbour sorts $\Vs(a)$ of a given sort $a$ is the set of components that can interact with it through actions of priority $1$.
  The set of neighbour actions $\Vh(a)$ of $a$ is the set of actions of priority $1$ influencing $a$.
  $$\Vs(a) = \{ b \in \components \mid \PHh(b, a) \neq \emptyset \}$$
  $$\Vh(a) = \{ h \in \PHh^{(1)} \mid \PHsort(\PHhitter(h)) \in \Vs(a) \wedge \PHsort(\PHtarget(h)) \in \Vs(a) \cup \{ a \} \}$$
\end{definition}

@@

In \pref{def:cs} we define the notion of well-formed cooperative sort.
A cooperative sort $a$ is well-formed iff
it is hit only by actions of priority $1$,
any sub-state of its neighbourhood is represented by exactly one local steady state in this sort,
and all of its processes represents at least one sub-state of its neighbourhood.
\begin{definition}
\label{def:cs}
  A sort $a \in \PHs$ is said to be a well-formed cooperative sort iff:
  \begin{itemize}
    \item $\forall h \in \PHh, \PHsort(\PHtarget(h)) = a \Rightarrow \prio(h) = 1$
    \item $\forall s \in \underset{b \in \Vs(a)}{\times} \PHl_b, |\pfp_s(a)| = 1$
    \item $\forall a_i \in \PHl_a, \exists s \in \underset{b \in \Vs(a)}{\times} \PHl_b, \pfp_s(a) = \{ a_i \}$
  \end{itemize}
\end{definition}
We call $\cs$ the set of well-formed cooperative sorts.
%We note that: $\forall a \in \components, \Vs(a) = \emptyset \wedge \Vh(a) = \emptyset$
As any local state of neighbour sorts of a cooperative sort leads to a local steady-state, we denote $\csState(a_i)$ the set of sub-states represented by the process $a_i$ of the cooperative sort $a$.
\begin{definition}[$\csState$]
\label{def:csState}
  \towrite{À reformuler}
  $$\csState(a_i) = \{ s \in \underset{b \in \Vs(a)}{\times} \PHl_b \mid \pfp_{s}(a) = a_i \}$$
\end{definition}

@@

\subsection{Consequences of the restrictions}
Given these hypotheses, we can denote: $\pfp_s(a) = a_i$ the local steady-state of $a$ in state $s$.
Furthermore, we denote by $\update(s)$ the state in which all cooperative sorts are updated.
\begin{align*}
  \update(s) = s \underset{a \in CS}{\Cap } \pfp_{s}(a)
\end{align*}
where for all process $a_i$, $s \Cap a_i$ is the state defined by:
$\PHget{(s \Cap a_i)}{a} = a_i \wedge \forall b \neq a, \PHget{(s \Cap a_i)}{b} = \PHget{s}{b}$.

\pref{th:update} states that from any state, there exists a scenario updating the cooperative sorts of this state.
\begin{theorem}
\label{th:update}
  $\forall s \in \PHl, \exists \delta \in \restriction{\Sce}{1}, s \PHplay \delta = \update(s)$
\end{theorem}
\begin{proof}
  Let $a$ be a cooperative sort so that $\PHget{s}{a} \neq \pfp_s(a)$.
  Given the definition of $\pfp_s(a)$, there exists a scenario updating $a$ in $s$.
  As there is no cycle in between the cooperative sorts and an updated cooperative sort cannot evolve, at most $|\cs|$ updates can be done.
\end{proof}
\pref{th:hcompcomp} states that for a given state $s$ and for any action $h = \PHhit{a_i}{b_j}{b_k}$ between well-formed components, $h$ can be played after a series of hits that do not prevent it to be fired.
\pref{th:hcscomp} states the same if $a$ is a well-formed cooperative sort, under the condition that $a_i$ reflects the state of the neighborhood of $a$.
\begin{theorem}
\label{th:hcompcomp}
  $\forall s \in \PHl, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh, a,b \in \components,$\\
  $(\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j) \Rightarrow (\exists \delta \in \restriction{\Sce}{1}, \text{$h$ can be played in $s \PHplay \delta$})$
\end{theorem}
\begin{proof}
  From \pref{th:update}, there exists a scenario $\delta$ so that: $(s \PHplay \delta) = \update(s)$.
  As $a,b \in \components$, $a_i \in s \PHplay \delta$ and $b_j \in s \PHplay \delta$.
  Finally, by definition of $\update(s)$, no action of priority $1$ can be played in $s \PHplay \delta$, thus $h$ can be played in $s \PHplay \delta$.
\end{proof}
\begin{theorem}
\label{th:hcscomp}
  $\forall s \in \PHl, \forall h = \PHhit{a_i}{b_j}{b_k} \in \PHh, a \in \cs, b \in \components$\\
  $(\PHget{s}{a} = a_i \wedge \PHget{s}{b} = b_j \wedge \pfp_s(a) = a_i) \Rightarrow (\exists \delta \in \restriction{\Sce}{1}, \text{$h$ can be played in $s \PHplay \delta$})$
\end{theorem}
\begin{proof}
  From \pref{th:update}, there exists a scenario $\delta$ so that: $(s \PHplay \delta) = \update(s)$.
  As $a_i \in \pfp_s(a)$, $a_i \in s \PHplay \delta$.
  Furthermore, $b \in \components$, thus $b_j \in s \PHplay \delta$.
  Finally, by definition of $\update(s)$, no action of priority $1$ can be played in $s \PHplay \delta$, thus $h$ can be played in $s \PHplay \delta$.
\end{proof}

### 2013-03-29 parts/sa.tex
defs + demo

Processus rencontrés :
\todo{Revenir à la définition simple de procs ?}
\begin{align*}
&\allprocs((\cwSol,\cwReq,\cwCont,\cwSat)) = \{ p \in \PHproc \mid \exists (P,ps) \in \cwSol, p \in ps \\
  &\qquad\qquad \vee p = \PHtarget(P) \vee \exists h \in \BS(P), (p = \PHhitter(h) \vee p = \PHbounce(h)) \\
  &\qquad\qquad \vee \exists(Q, \PHobj{p}{\PHbounce(Q)} \in \cwPrioCont \}
\end{align*}

@@

Mise à jour de l'état initial :
\begin{align*}
  \update(\ctx) = \ctx \Cap \bigcup_{a \in CS} \pfp_{\restriction{\ctx}{\Vs(a) \cup \{ a \}}}(a)
\end{align*}

@@

Let $(a_i, ps) \in \Brsp$ be a cooperative sort requirement whom all children are concretizable.
We label all processes of $ps$ by an integer: $ps = \{ p_n \}_{n \in \indexes{ps}}$.
Let us prove that for all $n \in \indexes{ps}$, there exists a scenario $\delta_n$ so that:
$\forall i \in \segm{1}{n}, \PHget{(s \PHplay \delta_n)}{\PHsort(p_i)} = p_i$.
\begin{itemize}
  \item It is straightforward for $\delta_0 = \varepsilon$.
  \item Let us suppose such $\delta_n$ exists. \todo{Il faut être plus précis pour prendre en compte les coopérations en cascade}
    By hypothesis, there exists $\delta' \in \muconcr_{s \PHplay \delta_n}(\PHobj{\any}{p_n})$.
%    so that: $\PHget{(s \PHplay \delta_n \PHplay \delta')}{\PHsort(p_{n+1})} = p_{n+1}$.
    By hypothesis, $(a_i, ps)$ is coherent (\pref{def:coherent}), which means that solving $p_{n+1}$ does not disturb the processes of $ps$.
    Therefore, $\forall i \in \segm{1}{n+1}$, $\PHget{(s \PHplay \delta_n \PHplay \delta')}{\PHsort(p_{i})} = p_{i}$.
    Finally, by theorem \todo{théorème mise à jour}, there exists a scenario $\delta'' \in \restriction{\Sce}{1}$
    that can be played in $s \PHplay \delta_n \PHplay \delta'$
    and so that, if we denote $\delta_{n+1} = \delta_n \PHplay \delta' \PHplay \delta''$,
    we have: $\update(s \PHplay \delta_n \PHplay \delta') = s \PHplay \delta_{n+1}$ (\todo{théorème stabilité}), and the same property about processes.
\end{itemize}
Therefore, $\delta = \delta_{\indexes{ps}}$ exists, and given its properties, we have: $\PHget{(s \PHplay \delta)}{a} = a_i$
and $\update(s \PHplay \delta) = s \PHplay \delta$.

@@

  \item Suppose all children objectives of $P$ are concretizable.
  \begin{itemize}
    \item If $\exists (P, Q) \in \Bcont$, then by hypothesis,
      $\muconcr_{s}(\obj{\PHtarget(P)}{\PHtarget(Q)} \concat Q) \neq \emptyset$, thus
      $\muconcr_{s}(P) \neq \emptyset$.
    \item Else, by \pref{def:maxCont}, the concretizations of the children of $P$ require no process of sort $\PHsort(P)$.
      Furthermore, there exists $\zeta \in \BS(P)$ so that $(P, \aZ) \in \Bsol$.
      We show by induction that for all $n \in \indexes{\zeta}$, there is a scenario $\delta_n$ so that $\PHget{(s \PHplay \delta_n)}{\PHsort(P)} = \PHbounce(\zeta_n)$.
%      We build recursively a scenario $\delta$. Let $m = |\indexes{\zeta}|$.
      \begin{itemize}
        \item[*] \towrite{construire récurvivement ? Distinguer les cas $b_j$ dans $\components$ ou $\cs$ (stabilité requise et obtenue)}
        Suppose that $\delta_n$ exists and let $\zeta_n = \PHhit{b_i}{a_j}{a_k}$.
        By hypothesis there exists $\delta' \in \muconcr_{s \PHplay \delta_n}(\PHobj{\any}{b_i})$ with $\PHsort(P) \notin \PHsort(\delta')$ by \pref{def:maxCont}.
        By \todo{théorème mise à jour} there exists $\delta'' \in \restriction{\Sce}{1}$ so that $\update(s \PHplay \delta') = s \PHplay \delta' \PHplay \delta''$.
        Furthermore, $\PHget{(s \PHplay \delta' \PHplay \delta'')}{b} = b_j$ \todo{car $b \in \components$ ou $b$ est stable}.
        Therefore, $\delta_{n+1} = \delta_n \PHplay \delta' \PHplay \delta'' \PHplay \zeta_n$.
      \end{itemize}
      Thus, $\delta_{\indexes{\zeta}} \in \muconcr_s(P)$. % and $\ceil(\delta) \subseteq max\ctx$.
  \end{itemize}
\end{itemize}

Finally, as $\muconcr_{max\ctx}(\w) \neq \emptyset$, $\uconcr(\w) \neq \emptyset$ (\pref{th:uconcr-ctx}).
\end{proof}

>>>>

The set of all processes involved in an abstract structure $\cwB$ is given in \pref{def:allprocs}.
\begin{definition}
\label{def:allprocs}
  \begin{align*}
  &\allprocs((\cwSol,\cwReq,\cwRSP,\cwCont)) = \{ p \in \PHproc \mid \exists (P,ps) \in \cwSol, \\
    &\qquad\qquad p \in ps \vee p = \PHtarget(P) \vee p = \PHbounce(P) \} \\
  %  &\qquad\qquad \vee p = \PHtarget(P) \vee \exists h \in \BS(P), (p = \PHhitter(h) \vee p = \PHbounce(h)) \\
  %  &\qquad\qquad \vee \exists(Q, \PHobj{p}{\PHbounce(Q)} \in \cwPrioCont \}
  \end{align*}
\end{definition}

@@

@@

Let $(a_i, ps) \in \Brsp$ be a cooperative sort requirement whom all children are concretizable.
We label all processes of $ps$ by an integer: $ps = \{ p_n \}_{n \in \indexes{ps}}$.
Let us prove that for all $n \in \indexes{ps}$, there exists a scenario $\delta_n$ so that:
$\forall i \in \segm{1}{n}, \PHget{(s \PHplay \delta_n)}{\PHsort(p_i)} = p_i$.
\begin{itemize}
  \item It is straightforward for $\delta_0 = \varepsilon$.
  \item Let us suppose such $\delta_n$ exists. \todo{Il faut être plus précis pour prendre en compte les coopérations en cascade}
    By hypothesis, there exists $\delta' \in \muconcr_{s \PHplay \delta_n}(\PHobj{\any}{p_n})$.
%    so that: $\PHget{(s \PHplay \delta_n \PHplay \delta')}{\PHsort(p_{n+1})} = p_{n+1}$.
    By hypothesis, $(a_i, ps)$ is coherent (\pref{def:coherent}), which means that solving $p_{n+1}$ does not disturb the processes of $ps$.
    Therefore, $\forall i \in \segm{1}{n+1}$, $\PHget{(s \PHplay \delta_n \PHplay \delta')}{\PHsort(p_{i})} = p_{i}$.
    Finally, by \pref{th:update}, there exists a scenario $\delta'' \in \restriction{\Sce}{1}$
    that can be played in $s \PHplay \delta_n \PHplay \delta'$
    and so that, if we denote $\delta_{n+1} = \delta_n \PHplay \delta' \PHplay \delta''$,
    we have: $\update(s \PHplay \delta_n \PHplay \delta') = s \PHplay \delta_{n+1}$ and the same property about processes (by \pref{th:hcscomp}).
\end{itemize}
Therefore, $\delta = \delta_{\indexes{ps}}$ exists, and given its properties, we have: $\PHget{(s \PHplay \delta)}{a} = a_i$
and $\update(s \PHplay \delta) = s \PHplay \delta$.

@@

  \item Suppose all children objectives of $P$ are concretizable.
  \begin{itemize}
    \item If $\exists (P, Q) \in \Bcont$, then by hypothesis,
      $\muconcr_{s}(\obj{\PHtarget(P)}{\PHtarget(Q)} \concat Q) \neq \emptyset$, thus
      $\muconcr_{s}(P) \neq \emptyset$.
    \item Else, by \pref{def:maxCont}, the concretizations of the children of $P$ require no process of sort $\PHsort(P)$.
      Furthermore, there exists $\zeta \in \BS(P)$ so that $(P, \aZ) \in \Bsol$.
      We show by induction that for all $n \in \indexes{\zeta}$, there is a scenario $\delta_n$ so that $\PHget{(s \PHplay \delta_n)}{\PHsort(P)} = \PHbounce(\zeta_n)$.
%      We build recursively a scenario $\delta$. Let $m = |\indexes{\zeta}|$.
      \begin{itemize}
        \item[*] \towrite{construire récurvivement ? Distinguer les cas $b_j$ dans $\components$ ou $\cs$ (stabilité requise et obtenue)}
        Suppose that $\delta_n$ exists and let $\zeta_n = \PHhit{b_i}{a_j}{a_k}$.
        By hypothesis there exists $\delta' \in \muconcr_{s \PHplay \delta_n}(\PHobj{\any}{b_i})$ with $\PHsort(P) \notin \PHsort(\delta')$ by \pref{def:maxCont}.
        By \pref{th:update} there exists $\delta'' \in \restriction{\Sce}{1}$ so that $\update(s \PHplay \delta') = s \PHplay \delta' \PHplay \delta''$.
        Furthermore, $\PHget{(s \PHplay \delta' \PHplay \delta'')}{b} = b_j$ (by \pref{th:hcompcomp} if $b \in \components$ or \pref{th:hcscomp} if $b \in \cs$).
        Therefore, $\delta_{n+1} = \delta_n \PHplay \delta' \PHplay \delta'' \PHplay \zeta_n$.
      \end{itemize}
      Thus, $\delta_{\indexes{\zeta}} \in \muconcr_s(P)$. % and $\ceil(\delta) \subseteq max\ctx$.
  \end{itemize}
\end{itemize}
Finally, as $\muconcr_{max\ctx}(\w) \neq \emptyset$, $\uconcr(\w) \neq \emptyset$ (\pref{th:uconcr-ctx}).
\end{proof}
